Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-16 draft.tex
Further revisions

Scatterplot visualization of all C5.0 classification results.

@@

The analysis of the dummy variables yields similar results like the \emph{blood pressure medication}.
%%
Variables, such as subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
%%
The classifier neglects the other group and yields a error below $0.5$.
%The dummy variable analysis yields fewer results.
%%
%It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
%%
%Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
%%
\subsubsection{Gender Groups}
%%
Classification using groups divided by gender do not produce satisfying results.
%%
Only \emph{hypothyroidism} could be described for male subjects with an error of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
%%
Since there are only 30 male subjects diagnosed with \emph{hypothyroidism}, the statistical power of the result is reduced.
%%
The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Body-Mass-Index Groups}
%%
Gender could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
%%
The result however is probably over fitted to the data due to tree sizes between $14$ and $16$.
%%

Some variables, such as \emph{body size} can be described with an error of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
%%
Notable is also the decreasing error with increasing decision tree size.
%%
Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
%%
The dichotomous variable whether the subject has felt pain in the last seven days can also be predicted for this group using the same features.
%%
The resulting tree consists of $8$ rules and has an error of $0.35$.
%%
Obese subjects are prone to back and leg pain due to a more stressed lumbar spine.
%%
The stress induced spine deformation seems to influence the pain levels for these subjects directly.
%How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.

@@

\subsubsection{Size Groups}
%%
Many previously described results are confounded with subject size.
%%
Differences between \emph{male} and \emph{female} subjects are explained by to the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
Back pain-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
%%
Radiating back pain could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, such as \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition} parameter, \emph{thyroid disorder} and \emph{social problems induced by back pain}.

@@

\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the parameters \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine.
%%
They are well suited to characterize subject gender due to their body size difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between \emph{males} and \emph{females}.
%%
Gender was characterized well in the analysis of all subjects as well as subjects divided by \emph{BMI}.

>>>>

Interactive scatterplot visualization of all C5.0 classification results.

@@

The analysis of the dummy variables yields similar results like the \emph{blood pressure medication}.
%%
Variables, such as subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
%%
The classifier neglects the other group and yields an error below $0.5$.
%The dummy variable analysis yields fewer results.
%%
%It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
%%
%Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
%%
\subsubsection{Gender Groups}
%%
Classification using groups divided by gender do not produce satisfying results.
%%
Only \emph{hypothyroidism} could be described for male subjects with an error of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
%%
Since there are only 30 male subjects diagnosed with \emph{hypothyroidism}, the statistical power of the result is reduced.
%%
The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Body-Mass-Index Groups}
%%
Gender could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
%%
The result however is probably overfitted to the data due to tree sizes between $14$ and $16$.
%%

Some variables, such as \emph{body size} can be described with an error of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
%%
%Notable is also the decreasing error with increasing decision tree size.
%%
Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
%%
The dichotomous variable whether the subject has felt pain in the last seven days can also be predicted for this group using the same features.
%%
The resulting tree consists of $8$ rules and has an error of $0.35$.
%%
Obese subjects are prone to back and leg pain due to a more stressed lumbar spine.
%%
The stress induced spine deformation seems to influence the pain levels for these subjects directly.
%How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.

@@

\subsubsection{Size Groups}
%%
Many previously described results are confounded with subject size.
%%
Differences between \emph{male} and \emph{female} subjects are explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameters or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
Back pain-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
%%
Radiating back pain could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition} parameter, \emph{thyroid disorder} and \emph{social problems induced by back pain}.

@@

\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the image based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine.
%%
They are well suited to characterize subject gender due to their body size difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between \emph{males} and \emph{females}.
%%
Gender was characterized well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
